An open-source full 3D electromagnetic modeler for 1D VTI media in Python: empymod

### 2017-05-31 ManuscriptGeophysics2016.tex
2nd Revision

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
\documentclass[paper]{geophysics}
% \documentclass[manuscript,revised]{geophysics}
%%fakesection ===    PACKAGES & DEFINITIONS    ===

@@

\title{An open-source \old{1D}\new{full 3D} electromagnetic modeler \new{for 1D VTI media} in Python: empymod}

@@

\footer{}
\lefthead{Werthmüller}
\righthead{Open-source\old{ 1D} EM modeler in Python}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
  The presented Python-code empymod computes the three-dimensional
  electromagnetic field in a layered-earth with vertical transverse isotropy by
  combining and extending two earlier presented algorithms in this journal.
  The bottleneck in frequency- and time-domain calculations of electromagnetic
  responses derived in the wavenumber-frequency domain is the transformations
  from wavenumber to space domain and from frequency to time domain, the
  so-called Hankel and Fourier transforms. Three different Hankel transform
  methods (quadrature, quadrature-with-extrapolation, and filters) and four
  different Fourier transform methods (FFT, FFTLog,
  quadrature-with-extrapolation, and filters) are included in empymod, which
  allows to compare these different methods in terms of speed and precision.
  The best transform in terms of speed and precision depends on the modeled
  frequencies: \old{available filters}\new{published digital filters for the
  Hankel transform}, for instance, are very fast and precise for frequencies in
  the range of controlled-source electromagnetic data, but fail in the
  frequency range of ground penetrating radar. Conventional quadrature, on the
  other hand, is in comparison very slow but can model any frequency.  Examples
  comparing empymod with analytical solutions and with existing electromagnetic
  modelers illustrate the capabilities of empymod.
\end{abstract}

@@

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It
\old{therefore} might \new{hence} prove to be valuable for students and
professionals alike, \new{and it is valuable for the objectives of reproducible
research.} It is written in Python, a modern, cross-platform, free and
open-source programming language. With its scientific libraries, mainly the
numeric and scientific modules \texttt{NumPy} and \texttt{SciPy}, it creates an
extremely powerful numerical calculation stack: Even though it is an
interpreted language it can be very fast, as \texttt{NumPy}/\texttt{SciPy} use
under the hood routines in Fortran or in C (using for instance the BLAS/Lapack
libraries) for the linear algebra computations, and Python is merely the glue.
The comparisons show that empymod is as fast as the existing codes. The code is
vectorized (avoiding loops) as much as possible, which improves computation.
The most time-consuming calculations have furthermore a flag to run
parallelized.  However, on a larger scale such as an inversion routine one
would probably want to parallelize the kernel calls instead of the kernel
itself. The code is hosted on GitHub, which makes it easy for anyone to improve
and contribute to the code\new{.}\old{: https://github.com/prisae/empymod}
\new{It comes with an extensive texting suite which is run automatically on
each commit, which guarantees some robustness and helps to avoid regression
bugs. Installation guidelines, documentation, and examples of its usage can be
found on its website at} \url{https://empymod.github.io}.%
\old{The documentation is hosted on https://empymod.readthedocs.io}
\cite{TLE.17.Werthmuller} gives a tutorial-style introduction to
\old{EM}\new{CSEM} modeling using empymod.

@@

The code consists of 5 files; these are 3 core modules plus \emph{utils}, which
contains input checks and other utilities, and \emph{filters}, containing the
FHT filter coefficients. The three core routines are: (1) \emph{kernel}, where
the wavenumber-domain calculation is carried out; (2) \emph{transform}, where
the Hankel and Fourier transforms are computed; and (3) \emph{model}, which
contains the actual modeling routines for end users. The main modeling routine
is \emph{bipole}, which can calculate frequency- and time-domain responses for
arbitrary oriented, electric or magnetic bipole sources and receivers of finite
length \new{(see also Appendix~\ref{apd:code})}.

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only. \old{Therefore}\new{Consequently}, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

@@

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module \old{numexpr}\new{\texttt{numexpr}}. This can significantly speed
up calculations if you run big models with many layers and for many offsets and
frequencies. However, if you include empymod in an inversion scheme then it
might be better to parallelize the calls to empymod, instead of empymod itself.

It is important to note that calculations in the wavenumber domain depend
\emph{only} on offset $r = \sqrt{x^2+y^2}$; the scaling factor, which depends
on the angle $\varphi = \rm{atan2}(y, x)$, is multiplied afterwards. In order
to calculate for instance a circle around the source, the kernel has to be
called only once \new{for offset $r$}, and subsequently scaled by the
angle-dependent factor for each source-receiver pair. This makes the splined
version very powerful, as it can be used for irregularly distributed data%
\old{ as long as all receiver depths are the same.}%
\new{: you can calculate any amount of data points in the x-y-plane in one
single kernel call. This advantage fails as soon as the receivers have
different depths.}

@@

\new{Comparisons of empymod to other modelers using non-homogeneous models
follow in the ground-penetrating radar example and in the section
\emph{Comparison to Key (2009, 2012)}.}

@@

Table~\ref{tbl:analytical} shows run times for these models. A few notes worth
mentioning: empymod carries out a lot of input checks, as it is quite forgiving
in what you input. EMmod, on the other hand, reads the input file and writes
the result to an output file, and copies the first quadrant to the other three.
Both have \old{therefore}\new{then} some overhead, and the comparison is not
strictly 1:1. For the comparison the same model was used as shown above, on a
regular grid with a spacing of 100\,m. On the test-machine the standard QWE and
FHT empymod would run into memory issues for denser spacing, hence arrays of
more than some 10\,000s of offsets. The splined and lagged versions of EMmod
could handle it, however.
%
\tabl[btp]{analytical}{Run times for the half-space model shown in Figure~1, on
  a regular grid with spacing of 100\,m, for EMmod and different Hankel
  transform settings of empymod (times in milliseconds).}{
  \centering
  \begin{tabularx}{320pt}{rAAABBB}
  \toprule%
  EMmod & \multicolumn{3}{c}{QWE} & \multicolumn{3}{c}{FHT} \\
        & 1 & 2 & 3               & 1 & 2 & 3 \\
  \midrule%
   5040 & 10300 & 2530 & 1210 & 1480 & 875 & 6 \\
  \bottomrule%
  \end{tabularx}
}
%
QWE becomes faster for decreasing precision, not surprisingly, and the splined
and lagged version of FHT are faster than the standard version. The lagged FHT
is the fastest by quite a margin, and about 1/3 of that time is from the input
checks, so it takes only a few milliseconds to calculate the 11\,025 offsets.
What is interesting here is that the lagged FHT is still more precise than
EMmod with the given settings, which means a speed-up of a factor 1000 for the
same result.

Many 1D CSEM codes use the diffusive approximation that is valid for low
frequencies. The appealing part of the derivation of \cite{GEO.15.Hunziker} is
that it models the complete wavefield, hence it is also valid for high
frequencies and therefore wave-phenomena. As an extreme case,
\cite{GEO.15.Hunziker} show a 1D example for ground-penetrating radar with a
center frequency of 250\,MHz. To do so, they calculate 2048 frequencies in the
range of 1\,MHz to 2048\,MHz for the Fast Fourier transform from frequency to
time domain. Here I only calculated the FFT with 850 frequencies from 1\,MHz
to 850\,MHz and then zero-padded up to 2048\,MHz, for EMmod and empymod; tests
have shown that this is sufficient. Figure~\ref{fig:gpr} shows the model with
the survey setup and the four results for EMmod, \new{empymod with }QUAD for
relative and absolute tolerance of $10^{-12}$ and $10^{-20}$, respectively,
empymod with a 21\,pt QWE and relative and absolute tolerance of $10^{-10}$ and
$10^{-18}$, respectively, and empymod with FHT with the filter Key09-401,
together with the analytical solution for the arrival of the direct wave (red),
the wave refracted at the surface (yellow), and the wave reflected at the
subsurface interface (green).

EMmod and the QUAD and QWE transforms of empymod yield pretty much the same
result. It is only at later times that some noise can be seen, different in
each method. FHT shows the first arrivals well, however, afterwards the result
becomes extremely noisy.
%
\plot*[btp]{gpr}{width=1\textwidth}{GPR example for (b) EMmod, (c) QUAD
  for relative and absolute tolerance of $10^{-12}$ and $10^{-20}$,
  respectively, (d) empymod with 51\,pt QWE for relative and absolute tolerance
  of $10^{-8}$ and $10^{-15}$, respectively, and (e) empymod with FHT
  (Key09-401); run times are given in the subplot titles.}
%
Interesting are the calculation times for these three results. EMmod took
roughly 9.5 hours to calculate, empymod with QUAD about 7.8 hours, empymod with
QWE about 4.4 hours, and empymod with FHT under a minute. Note that QWE has
internally a check: If the kernel is too steep within given intervals, it uses
QUAD instead, as QWE cannot handle very steep functions. QWE uses in about 1/3
% (37%)
of the calculation of this example QUAD. Surprising is how good the FHT result
is, given that the filter was designed for CSEM data, hence frequencies 6 to 9
orders of magnitudes smaller and much bigger offsets. In any case,
EMmod/empymod are not optimized for nor intended to model GPR data, and
calculations in time-domain will be much faster and more precise. It is
nevertheless an interesting proof of concept and shows that EMmod/empymod model
the entire EM field. This leaves the question if it would be possible to
create a filter that works for the frequencies and the offsets required for GPR
calculations.

\new{Note that a common-midpoint (CMP) GPR-layout was chosen with increasing
offsets between the antennas, rather than the more common profiling layout with
fixed antenna separation, often with shielded antennas. The reason is that a
fixed offset calculation with shielded antennas of a 1D model would result in
exactly the same response at each measurement point, resulting in two flat
horizons, the direct and the reflected field. A CMP example is for this reason
more interesting than a profiling example.}

@@

Outside of the traditional scope empymod can be very educative for someone who
is just starting to get interested in electromagnetic modeling or even just in
Hankel and Fourier transforms, hence for educational purpose: it is well suited
for students, specifically as Python is becoming more and more widespread in
academia, and the number of Universities that teach geophysicist Python (mostly
instead of Matlab) is growing annually. Python, like Matlab, has the advantage
of very fast developing times, and considerably lower entry barriers than C or
Fortran for beginners. \new{And, unlike Matlab, Python is completely free and
open-source. The module empymod is accordingly ideal for truly reproducible
research.}

@@

\new{
Figure~\ref{fig:codestructure} shows the structure of empymod.%
}
%
\plot*[btp]{codestructure}{}{Code structure of the module empymod showing the
  different files and the main modeling routines \texttt{bipole},
  \texttt{dipole}, and \texttt{analytical}.}%
%
\new{
The important routines for a user who only wants to use the code and is not
interested in the details are located in \texttt{model.py}. The three main
modeling routines are \texttt{bipole}, \texttt{dipole}, and
\texttt{analytical}. With \texttt{bipole} you can calculate frequency- and
time-domain responses for arbitrary oriented, electric or magnetic bipole
sources and receivers of finite length. The routine \texttt{dipole} is
basically a subset of \texttt{bipole}, and is a good starting point if you
want to create your own modeling routine. And \texttt{analytical} is an
interface to the analytical full- and halfspace solutions that come with
empymod. In addition there are the routines \texttt{wavenumber} to calculate
the wavenumber-frequency result of a dipole, and \texttt{gpr} to calculate the
response for a ground-penetrating radar with a given center frequency. More
modeling routines will hopefully be added in the future.
}

\new{
The file \texttt{kernel.py} contains the wavenumber-frequency domain
calculation as given in \cite{GEO.16.Hunziker}. It also contains analytical
solutions, namely the complete space-frequency domain fullspace solution as
given in \cite{GEO.16.Hunziker}, and the space-frequency and space-time domain
diffusive halfspace solutions for electric sources and receivers as given in
\cite{PIER.10.Slob}. All calculations related to the Hankel and Fourier
transforms are carried out in the file \texttt{transform.py}.  Digital filters
are collected in \texttt{filter.py}. It contains, as of now, 9 filters for the
Hankel transform \citep{TMS.82.Anderson, GEO.12.Key, GEO.09.Key, GP.07.Kong},
and 5 filters for the Fourier transform \citep{GEO.12.Key, GEO.09.Key}.
}

\new{
The modeling routines rely heavily on \texttt{utils.py}. In this routine are
the input tests and other checks, such as to get the layer number in which
source and receivers reside, and everything related to the verbosity of the
code.  And \texttt{\_\_init\_\_.py} is a required file, such that Python
recognises empymod as a module that can be loaded. In the case of empymod it
does three more things, common to Python init-files: It provides the version
number, it loads the most important routine (\texttt{bipole}, \texttt{dipole},
\texttt{analytical}) so they are available from the module itself, and it
contains some documentation.
}

\new{
As an exemplary minimal example we calculate a halfspace-model with
air/halfspace interface at $0\,$m; electric, $x$-directed source at
$x_s=y_s=0\,$m and $z_s=1\,$m; electric receiver at $x_r=1000\,$m, $y_r=0\,$m,
$z_r=1\,$m, with azimuth $\theta_r=45^\circ$ and dip $\varphi_r=10^\circ$; and
frequency of $2\,$Hz.
}

@@

\new{The function \texttt{bipole} takes many more input arguments, but these
five are mandatory. You can find more extensive examples under
\url{https://empymod.github.io} or in the Geophysical Tutorial of
\cite{TLE.17.Werthmuller}.}

>>>>

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
\documentclass[manuscript,revised]{geophysics}
%%fakesection ===    PACKAGES & DEFINITIONS    ===

@@

\title{An open-source full 3D electromagnetic modeler for 1D VTI media in Python: empymod}

@@

\footer{}
\lefthead{Werthmüller}
\righthead{Open-source EM modeler in Python}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
  The presented Python-code empymod computes the three-dimensional
  electromagnetic field in a layered-earth with vertical transverse isotropy by
  combining and extending two earlier presented algorithms in this journal.
  The bottleneck in frequency- and time-domain calculations of electromagnetic
  responses derived in the wavenumber-frequency domain is the transformations
  from wavenumber to space domain and from frequency to time domain, the
  so-called Hankel and Fourier transforms. Three different Hankel transform
  methods (quadrature, quadrature-with-extrapolation, and filters) and four
  different Fourier transform methods (FFT, FFTLog,
  quadrature-with-extrapolation, and filters) are included in empymod, which
  allows to compare these different methods in terms of speed and precision.
  The best transform in terms of speed and precision depends on the modeled
  frequencies: published digital filters for the Hankel transform, for
  instance, are very fast and precise for frequencies in the range of
  controlled-source electromagnetic data, but fail in the frequency range of
  ground penetrating radar. Conventional quadrature, on the other hand, is in
  comparison very slow but can model any frequency.  Examples comparing empymod
  with analytical solutions and with existing electromagnetic modelers
  illustrate the capabilities of empymod.
\end{abstract}

@@

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It might
hence prove to be valuable for students and professionals alike, and it is
valuable for the objectives of reproducible research. It is written in Python,
a modern, cross-platform, free and open-source programming language. With its
scientific libraries, mainly the numeric and scientific modules \texttt{NumPy}
and \texttt{SciPy}, it creates an extremely powerful numerical calculation
stack: Even though it is an interpreted language it can be very fast, as
\texttt{NumPy}/\texttt{SciPy} use under the hood routines in Fortran or in C
(using for instance the BLAS/Lapack libraries) for the linear algebra
computations, and Python is merely the glue.  The comparisons show that empymod
is as fast as the existing codes. The code is vectorized (avoiding loops) as
much as possible, which improves computation.  The most time-consuming
calculations have furthermore a flag to run parallelized.  However, on a larger
scale such as an inversion routine one would probably want to parallelize the
kernel calls instead of the kernel itself. The code is hosted on GitHub, which
makes it easy for anyone to improve and contribute to the code.  It comes with
an extensive testing suite which is run automatically on each commit, which
guarantees some robustness and helps to avoid regression bugs. Installation
guidelines, documentation, and examples of its usage can be found on its
website at \url{https://empymod.github.io}.  \cite{TLE.17.Werthmuller} gives a
tutorial-style introduction to CSEM modeling using empymod.

@@

The code consists of 5 files; these are 3 core modules plus \emph{utils}, which
contains input checks and other utilities, and \emph{filters}, containing the
FHT filter coefficients. The three core routines are: (1) \emph{kernel}, where
the wavenumber-domain calculation is carried out; (2) \emph{transform}, where
the Hankel and Fourier transforms are computed; and (3) \emph{model}, which
contains the actual modeling routines for end users. The main modeling routine
is \emph{bipole}, which can calculate frequency- and time-domain responses for
arbitrary oriented, electric or magnetic bipole sources and receivers of finite
length (see also Appendix~\ref{apd:code}).

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only. Consequently, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

@@

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module \texttt{numexpr}. This can significantly speed up calculations if
you run big models with many layers and for many offsets and frequencies.
However, if you include empymod in an inversion scheme then it might be better
to parallelize the calls to empymod, instead of empymod itself.

It is important to note that calculations in the wavenumber domain depend
\emph{only} on offset $r = \sqrt{x^2+y^2}$; the scaling factor, which depends
on the angle $\varphi = \rm{atan2}(y, x)$, is multiplied afterwards. In order
to calculate for instance a circle around the source, the kernel has to be
called only once for offset $r$, and subsequently scaled by the
angle-dependent factor for each source-receiver pair. This makes the splined
version very powerful, as it can be used for irregularly distributed data: you
can calculate any amount of data points in the x-y-plane in one single kernel
call. This advantage fails as soon as the receivers have different depths.

@@

Comparisons of empymod to other modelers using non-homogeneous models
follow in the ground-penetrating radar example and in the section
\emph{Comparison to Key (2009, 2012)}.

@@

Table~\ref{tbl:analytical} shows run times for these models. A few notes worth
mentioning: empymod carries out a lot of input checks, as it is quite forgiving
in what you input. EMmod, on the other hand, reads the input file and writes
the result to an output file, and copies the first quadrant to the other three.
Both have then some overhead, and the comparison is not strictly 1:1. For the
comparison the same model was used as shown above, on a regular grid with a
spacing of 100\,m. On the test-machine the standard QWE and FHT empymod would
run into memory issues for denser spacing, hence arrays of more than some
10\,000s of offsets. The splined and lagged versions of EMmod could handle it,
however.
%
\tabl[btp]{analytical}{Run times for the half-space model shown in Figure~1, on
  a regular grid with spacing of 100\,m, for EMmod and different Hankel
  transform settings of empymod (times in milliseconds).}{
  \centering
  \begin{tabularx}{320pt}{rAAABBB}
  \toprule%
  EMmod & \multicolumn{3}{c}{QWE} & \multicolumn{3}{c}{FHT} \\
        & 1 & 2 & 3               & 1 & 2 & 3 \\
  \midrule%
   5040 & 10300 & 2530 & 1210 & 1480 & 875 & 6 \\
  \bottomrule%
  \end{tabularx}
}
%
QWE becomes faster for decreasing precision, not surprisingly, and the splined
and lagged version of FHT are faster than the standard version. The lagged FHT
is the fastest by quite a margin, and about 1/3 of that time is from the input
checks, so it takes only a few milliseconds to calculate the 11\,025 offsets.
What is interesting here is that the lagged FHT is still more precise than
EMmod with the given settings, which means a speed-up of a factor 1000 for the
same result.

Many 1D CSEM codes use the diffusive approximation that is valid for low
frequencies. The appealing part of the derivation of \cite{GEO.15.Hunziker} is
that it models the complete wavefield, hence it is also valid for high
frequencies and therefore wave-phenomena. As an extreme case,
\cite{GEO.15.Hunziker} show a 1D example for ground-penetrating radar with a
center frequency of 250\,MHz. To do so, they calculate 2048 frequencies in the
range of 1\,MHz to 2048\,MHz for the Fast Fourier transform from frequency to
time domain. Here I only calculated the FFT with 850 frequencies from 1\,MHz
to 850\,MHz and then zero-padded up to 2048\,MHz, for EMmod and empymod; tests
have shown that this is sufficient. Figure~\ref{fig:gpr} shows the model with
the survey setup and the four results for EMmod, empymod with QUAD for
relative and absolute tolerance of $10^{-12}$ and $10^{-20}$, respectively,
empymod with a 21\,pt QWE and relative and absolute tolerance of $10^{-10}$ and
$10^{-18}$, respectively, and empymod with FHT with the filter Key09-401,
together with the analytical solution for the arrival of the direct wave (red),
the wave refracted at the surface (yellow), and the wave reflected at the
subsurface interface (green).

EMmod and the QUAD and QWE transforms of empymod yield pretty much the same
result. It is only at later times that some noise can be seen, different in
each method. FHT shows the first arrivals well, however, afterwards the result
becomes extremely noisy.
%
\plot*[btp]{gpr}{width=1\textwidth}{GPR example for (b) EMmod, (c) empymod
with QUAD for relative and absolute tolerance of $10^{-12}$ and $10^{-20}$,
respectively, (d) empymod with 51\,pt QWE for relative and absolute tolerance
of $10^{-8}$ and $10^{-15}$, respectively, and (e) empymod with FHT
(Key09-401); run times are given in the subplot titles.}
%
Interesting are the calculation times for these three results. EMmod took
roughly 9.5 hours to calculate, empymod with QUAD about 7.8 hours, empymod with
QWE about 4.4 hours, and empymod with FHT under a minute. Note that QWE has
internally a check: If the kernel is too steep within given intervals, it uses
QUAD instead, as QWE cannot handle very steep functions. QWE uses in about 1/3
% (37%)
of the calculation of this example QUAD. Surprising is how good the FHT result
is, given that the filter was designed for CSEM data, hence frequencies 6 to 9
orders of magnitudes smaller and much bigger offsets. In any case,
EMmod/empymod are not optimized for nor intended to model GPR data, and
calculations in time-domain will be much faster and more precise. It is
nevertheless an interesting proof of concept and shows that EMmod/empymod model
the entire EM field. This leaves the question if it would be possible to
create a filter that works for the frequencies and the offsets required for GPR
calculations.

Note that a common-midpoint (CMP) GPR-layout was chosen with increasing
offsets between the antennas, rather than the more common profiling layout with
fixed antenna separation, often with shielded antennas. The reason is that a
fixed offset calculation with shielded antennas of a 1D model would result in
exactly the same response at each measurement point, resulting in two flat
horizons, the direct and the reflected field. A CMP example is for this reason
more interesting than a profiling example.

@@

Outside of the traditional scope empymod can be very educative for someone who
is just starting to get interested in electromagnetic modeling or even just in
Hankel and Fourier transforms, hence for educational purpose: it is well suited
for students, specifically as Python is becoming more and more widespread in
academia, and the number of Universities that teach geophysicist Python (mostly
instead of Matlab) is growing annually. Python, like Matlab, has the advantage
of very fast developing times, and considerably lower entry barriers than C or
Fortran for beginners. And, unlike Matlab, Python is completely free and
open-source. The module empymod is accordingly ideal for truly reproducible
research.

@@

Figure~\ref{fig:codestructure} shows the structure of empymod.%
%
\plot*[btp]{codestructure}{}{Code structure of the module empymod showing
  the different files and the main modeling routines \texttt{bipole},
  \texttt{dipole}, and \texttt{analytical}.}%
%
The important routines for a user who only wants to use the code and is not
interested in the details are located in \texttt{model.py}. The three main
modeling routines are \texttt{bipole}, \texttt{dipole}, and
\texttt{analytical}. With \texttt{bipole} you can calculate frequency- and
time-domain responses for arbitrary oriented, electric or magnetic bipole
sources and receivers of finite length. The routine \texttt{dipole} is
basically a subset of \texttt{bipole}, and is a good starting point if you
want to create your own modeling routine. And \texttt{analytical} is an
interface to the analytical full- and halfspace solutions that come with
empymod. In addition there are the routines \texttt{wavenumber} to calculate
the wavenumber-frequency result of a dipole, and \texttt{gpr} to calculate the
response for a ground-penetrating radar with a given center frequency. More
modeling routines will hopefully be added in the future.

The file \texttt{kernel.py} contains the wavenumber-frequency domain
calculation as given in \cite{GEO.16.Hunziker}. It also contains analytical
solutions, namely the complete space-frequency domain fullspace solution as
given in \cite{GEO.16.Hunziker}, and the space-frequency and space-time domain
diffusive halfspace solutions for electric sources and receivers as given in
\cite{PIER.10.Slob}. All calculations related to the Hankel and Fourier
transforms are carried out in the file \texttt{transform.py}.  Digital filters
are collected in \texttt{filter.py}. It contains, as of now, 9 filters for the
Hankel transform \citep{TMS.82.Anderson, GEO.12.Key, GEO.09.Key, GP.07.Kong},
and 5 filters for the Fourier transform \citep{GEO.12.Key, GEO.09.Key}.

The modeling routines rely heavily on \texttt{utils.py}. In this routine are
the input tests and other checks, as well as everything related to the
verbosity of the code.  And \texttt{\_\_init\_\_.py} is a required file, such
that Python recognises empymod as a module that can be loaded. In the case of
empymod it does three more things, common to Python init-files: It provides the
version number, it loads the most important routine (\texttt{bipole},
\texttt{dipole}, \texttt{analytical}) so they are available from the module
itself, and it contains some documentation.

As an exemplary minimal example we calculate a halfspace-model with
air/halfspace interface at $0\,$m; electric, $x$-directed source at
$x_s=y_s=0\,$m and $z_s=1\,$m; electric receiver at $x_r=1000\,$m, $y_r=0\,$m,
$z_r=1\,$m, with azimuth $\theta_r=45^\circ$ and dip $\varphi_r=10^\circ$; and
frequency of $2\,$Hz.

@@

The function \texttt{bipole} takes many more input arguments, but these
five are mandatory. You can find more extensive examples under
\url{https://empymod.github.io} or in the Geophysical Tutorial of
\cite{TLE.17.Werthmuller}.
